Vertical and oblique transmission fluctuating in time and in space

### 2018-06-17 ms/ms_sunnyvale.tex
UL comments

% Title page
\title{Vertical and oblique transmission with fluctuating transmission}

@@

Suppose that the vertical transmission rate $\rho$ fluctuates, such that $\rho = \rho_1$ in odd generations and $\rho = \rho_2$ in even generations.
From \eqref{eq:recurrence}, the recurrence equations for two generations is then
\begin{equation}\begin{aligned} \label{eq:recurrence_two_generations}
x' = \rho_1 \frac{1+s}{\overline w} x + (1-\rho_1)x, \\
x'' = \rho_2 \frac{1+s}{\overline w} x' + (1-\rho_2)x'.
\end{aligned}\end{equation}

@@

\paragraph{Symmetric periods---$AkBk$.}
We first consider environments that fluctuate periodically every $k$ generations between favoring $A$ and $B$.
Using a linear approximation, fixations of $A$ ($x^*=1$) and $B$ ($x^*=0$) are locally stable if, respectively,
\begin{equation}\begin{aligned}
\big[F'_A(1) F'_B(1)\big]^k < 1, \\
\big[F'_A(0) F'_B(0)\big]^k < 1. \\
\end{aligned}\end{equation}
A \emph{protected polymorphism} exists if neither of these conditions are met.
Therefore, for a protected polymorphism to exist, we require
\begin{equation}
1 < F'_A(1) F'_B(1) = \Big(1-\rho_A\frac{s}{1+s}\Big)\Big(1+\rho_B s\Big) 
= 1+\frac{s}{1+s}\big[\rho_B(1+s-s\rho_A)-\rho_A\big],
\end{equation}
and
\begin{equation}
1 < F'_A(0) F'_B(0) = \Big(1+\rho_A s\Big)\Big(1-\rho_B \frac{s}{1+s}\Big) 
= 1+\frac{s}{1+s}\big[\rho_A(1+s) -\rho_B(1+\rho_A s)\big],
\end{equation}
which can be summarized as a condition on $\rho_B$,
\begin{equation} \label{eq:poly_condition_periodic_fluc}
\frac{\rho_A}{1+s(1-\rho_A)} < \rho_B < \frac{(1+s)\rho_A}{1+s\rho_A},
\end{equation}
or as a condition on the difference between $\rho_B$ and $\rho_A$
\begin{equation}
-s\rho_B(1-\rho_A) < \rho_B - \rho_A < s\rho_A(1-\rho_B).
\end{equation}
Surprisingly, the stronger the selection (i.e. the greater the value of $s$), the larger the difference in vertical transmission rates that allows a protected polymorphism.

@@

\paragraph{Asymmetric period---$AkBl$.}
More generally, phenotype $A$ could be favored for $k$ generations and phenotype $B$ for $l$ generations, with the transmission rate following the same cycle with $\rho=\rho_A$ when $A$ is favored and $\rho=\rho_B$ when $B$ is favored.
The requirements for a protected polymorphism are now
\begin{equation}\begin{aligned}
1 &< (F'_A(0))^k (F'_B(0))^l = a^k \big(b - \Delta \rho s/(1+s)\big)^l,\\
1 &< (F'_A(1))^k (F'_B(1))^l = b^k (a + \Delta \rho s)^l,
\end{aligned}\end{equation}
where $a=1+\rho_A s>1$, $b=1-\rho_A\frac{s}{1+s}<1$, and $\Delta \rho = \rho_B - \rho_A$.
This leads to a condition similar to eq.~20 in~\citet{Ram2018}, but more complex due to the inclusion of $\Delta \rho$:
\begin{equation} \label{eq:poly_condition_periodic_fluc_k_l}
\frac{-\log{b}}{\log{\big(a + s \Delta \rho \big)}} < 
\frac{l}{k} < 
\frac{\log{a}}{-\log{\big(b - s \Delta \rho/(1+s)\big)}}.
\end{equation}
Therefore, for a given value of $\rho_A$, if $\rho_B>\rho_A$ then increasing the vertical transmission rate of $B$ decreases the environmental period ratio $l/k$ that permits protected polymorphism; decreasing $\rho_B$  will have the opposite effect; it increases the ratio $l/k$ that permits a protected polymorphism.

@@

Now suppose that both selection and transmission fluctuate randomly.
Rewrite eq.~28 from \citet{Ram2018} so that $\rho$ is also a random variable
\begin{equation}
x_t = x_t \frac{1 + z_t + x_t (1 - \rho_t) s_t}{1 + x_t s_t},
\end{equation}
where $s_t$ are i.i.d (independent and identically distributed), $Pr(-1+C<s_t<D)=1$ for some positive $C$ and $D$, $\rho_t$ are i.i.d, and  $0<\rho_t<1$ ($t=0,1,2,\ldots$).
Therefore, $z_t = \rho_t s_t$  are independent and identically distributed and $P(-1+C < z_t < D)$. 
From \citet[][results~6 and 7]{Ram2018} we have:
\begin{itemize}
\item Suppose $E[log(1+z_t)]>0$. Then $x^*=0$ is not stochastically locally stable and in fact $P(lim_{t \to \infty} x_t=0) = 0$, i.e., fixation of $B$ almost surely does not occur.
\item Suppose $E[log(1+z_t)]<0$. Then $x^*=0$ is stochastically locally stable. 
\item Similarly, if $E[log(1-z_t/(1+s_t)]<0$, then $x^*=1$ is stochastically locally stable, and if $E[log(1-z_t/(1+s_t))]>0$, then fixation of $A$ almost surely does not occur.
\item In particular, if $E[z_t] = cov(\rho_t, s_t) + E[\rho_t] E[s_t] \le 0$ then $x^*=0$ is stochastically locally stable, and similarly if $E[-z_t/(1+s_t)] = cov(\rho_t, -s_t/(1+s_t)) - E[\rho_t] E[s_t/(1+s_t)] \le 0$ then $x^*=1$ is stochastically locally stable.
\item It is not possible that $E[log(1+z_t)]$ and $E[log(1-z_t/(1+s_t)]$ are both negative, as their sum is positive
\begin{multline}
E[log(1+z_t)] + E[log(1-z_t/(1+s_t)] = \\
E[log(1+z_t) + log(1-z_t/(1+s_t)]= \\
E[log\big((1+z_t)(1-z_t/(1+s_t)\big)]= \\
E[log\big( 1+\rho_t(1-\rho_t)s_t^2/(1+s_t) \big)] > 0,
\end{multline}
and therefore it is not possible that both fixations are stochastically locally stable.
\end{itemize}

\paragraph{Examples.}
First, if $s_t$ and $\rho_t$ are independent ($cov(s_t, \rho_t)=0$) and $s_t$ is symmetric around zero, then $E[z_t]=0$ and $E[-z_t/(1+s_t)]>0$ (because $E[s_t/(1+s_t)] < E[s_t]$).
Therefore, fixation of $B$ is stochastically locally stable and fixation of $A$ almost surely does not occur.
For example, let $s_t \sim U(-1, 1)$ and $\rho_t \sim U(0,1)$ independently (in particular, $cov(\rho_t, s_t)=0$), then $E[log(1+z_t)]\approx -0.07315$ and $E[log(1-z_t/(1+s_t))]\approx 0.2337$.

@@

% TODO full correlation between rho and s?

@@

We obtain a result similar to result 11 from~\citet{Ram2018}:
The mean $\mu(x)$ and variance $\sigma^2(x)$ of the change in the frequency $x$ of phenotype $A$ in the case of a cycling environment $AkBl$, where $k+l=n$, are
\begin{equation} \label{eq:drift_diffusion_terms}
\mu(x) = S_n x(1-x)
\quad \text{and} \quad
\sigma^2(x) = n x (1-x),
\end{equation}
where $S_n = \sum_{t=1}^{n}{z_t}$.
Furthermore, combining \eqref{eq:drift_diffusion_terms} with eqs.~46-47 from \citet{Ram2018}, we find that the fixation probability of phenotype $A$, starting from a frequency $x$ of phenotype $A$, is
\begin{equation}
u(x) = \frac{1 - e^{-2 \frac{S_n}{n} x}}{1 - e^{-2 \frac{S_n}{n}}}.
\end{equation}
From result 10 in \citep{Ram2018}, $u(x)$ is monotone increasing in $S_n/n$ which is the average selection coefficient of $A$ weighted by the vertical transmission rates $\rho_t$.
Therefore, if $s_t$ and $\rho_t$ are positively (negatively) correlated, $S_n/n$ increases (decreases), and the fixation probability $u(x)$ increases (decreases). 
This occurs because selection only affects those individuals that transmit their phenotype to their own offspring (i.e. vertically), and a fraction $1-\rho_t$ of the population individuals is effectively masked (for better or worse) from selection at each generation.

@@

\autoref{fig:migration_rho} and \autoref{fig:migration_m} show the stable frequencies of phenotype $A$~ (eq.~\ref{eq:migration_model_unconditional_symmetric_polymorphism}) and the stable population mean fitnesses in the two demes with symmetric selection where $A$ is favored in deme 1 and $B$ is favored in deme 2 with similar selection intensities.
Notably, in the absence of oblique transmission (\autoref{fig:migration_m}, left column), migration decreases the differences between the demes and reduces the population mean fitnesses.
With some oblique transmission, but equal in both demes, results are similar (not shown).
However, if oblique transmission is stronger in deme 2 than in deme 1 (\autoref{fig:migration_m}, middle and right columns), the stable frequency of $A$ increases in both demes.
Therefore, the mean fitness in deme 1 decreases to a lesser extent than in deme 2, and even increases when the migration rate is high enough.

@@

\newpage

>>>>

% Title page
\title{Vertical and oblique transmission fluctuating in time and in space}

@@

Suppose that the vertical transmission rate $\rho$ fluctuates, such that $\rho = \rho_1$ in odd generations and $\rho = \rho_2$ in even generations.
From \eqref{eq:recurrence}, the recurrence equations for two generations is then
\begin{equation}\begin{aligned} \label{eq:recurrence_two_generations}
x' = \rho_1 \frac{1+s}{\overline w} x + (1-\rho_1)x, \quad \overline w = 1 + xs, \\
x'' = \rho_2 \frac{1+s}{\overline w'} x' + (1-\rho_2)x', \quad \overline w' = 1 + x's.
\end{aligned}\end{equation}

@@

\paragraph{Symmetric periods---$AkBk$.}
We first consider environments that fluctuate periodically every $k$ generations between favoring $A$ and $B$.
Using a linear approximation, fixations of $A$ ($x^*=1$) and $B$ ($x^*=0$) are locally stable if, respectively,
\begin{equation}\begin{aligned}
\big[F'_A(1) F'_B(1)\big]^k < 1, \\
\big[F'_A(0) F'_B(0)\big]^k < 1. \\
\end{aligned}\end{equation}
A \emph{protected polymorphism} exists if neither of these conditions are met.
Therefore, for a protected polymorphism to exist, we require
\begin{equation}
1 < F'_A(1) F'_B(1) = \Big(1-\rho_A\frac{s}{1+s}\Big)\Big(1+\rho_B s\Big) 
= 1+\frac{s}{1+s}\big[\rho_B(1+s-s\rho_A)-\rho_A\big],
\end{equation}
and
\begin{equation}
1 < F'_A(0) F'_B(0) = \Big(1+\rho_A s\Big)\Big(1-\rho_B \frac{s}{1+s}\Big) 
= 1+\frac{s}{1+s}\big[\rho_A(1+s-s\rho_B)-\rho_B\big],
\end{equation}
which can be summarized as a condition on $\rho_B$,
\begin{equation} \label{eq:poly_condition_periodic_fluc}
\frac{\rho_A}{1+s(1-\rho_A)} < \rho_B < \frac{(1+s)\rho_A}{1+s\rho_A},
\end{equation}
or as a condition on the difference between $\rho_B$ and $\rho_A$
\begin{equation}
-s\rho_B(1-\rho_A) < \rho_B - \rho_A < s\rho_A(1-\rho_B).
\end{equation}
Surprisingly, the stronger the selection (i.e. the greater the value of $s$), the larger the difference in vertical transmission rates that allows a protected polymorphism.

@@

\paragraph{Asymmetric period---$AkBl$.}
More generally, phenotype $A$ could be favored for $k$ generations and phenotype $B$ for $l$ generations, with the transmission rate following the same cycle with $\rho=\rho_A$ when $A$ is favored and $\rho=\rho_B$ when $B$ is favored.
The requirements for a protected polymorphism are now
\begin{equation}\begin{aligned}
1 &< (F'_A(0))^k (F'_B(0))^l = a^k \big(b - s \Delta \rho /(1+s)\big)^l,\\
1 &< (F'_A(1))^k (F'_B(1))^l = b^k (a + s \Delta \rho)^l,
\end{aligned}\end{equation}
where $a=1+\rho_A s>1$, $b=1-\rho_A\frac{s}{1+s}<1$, and $\Delta \rho = \rho_B - \rho_A$.
This leads to a condition similar to eq.~20 in~\citet{Ram2018}, but more complex due to the inclusion of $\Delta \rho$:
\begin{equation} \label{eq:poly_condition_periodic_fluc_k_l}
\frac{-\log{b}}{\log{\big(a + s \Delta \rho \big)}} < 
\frac{l}{k} < 
\frac{\log{a}}{-\log{\big(b - s \Delta \rho/(1+s)\big)}}.
\end{equation}
Therefore, for a given value of $\rho_A$, if $\rho_B>\rho_A$ then increasing the vertical transmission rate of $B$ decreases the environmental period ratio $l/k$ that permits protected polymorphism; decreasing $\rho_B$  will have the opposite effect; it increases the ratio $l/k$ that permits a protected polymorphism.

@@

Now suppose that both selection and transmission fluctuate randomly.
Rewrite eq.~28 from \citet{Ram2018} so that $\rho$ is also a random variable
\begin{equation}
x_t = x_t \frac{1 + z_t + x_t (1 - \rho_t) s_t}{1 + x_t s_t},
\end{equation}
where $z_t=\rho_t s_t$, $s_t$ are i.i.d (independent and identically distributed), $Pr(-1+C<s_t<D)=1$ for some positive $C$ and $D$, $\rho_t$ are i.i.d, and  $0<\rho_t<1$ ($t=0,1,2,\ldots$).
Therefore, $z_t$ are independent and identically distributed and $P(-1+C < z_t < D)$. 
From \citet[][results~6 and 7]{Ram2018} we have:
\begin{itemize}
\item Suppose $E[log(1+z_t)]>0$. Then $x^*=0$ is not stochastically locally stable and in fact $P(lim_{t \to \infty} x_t=0) = 0$, i.e., fixation of $B$ almost surely does not occur.
\item Suppose $E[log(1+z_t)]<0$. Then $x^*=0$ is stochastically locally stable. 
\item Similarly, if $E[log(1-z_t/(1+s_t)]<0$, then $x^*=1$ is stochastically locally stable, and if $E[log(1-z_t/(1+s_t))]>0$, then fixation of $A$ almost surely does not occur.
\item In particular, if $E[z_t] = cov(\rho_t, s_t) + E[\rho_t] E[s_t] \le 0$ then $x^*=0$ is stochastically locally stable, and similarly if $E[-z_t/(1+s_t)] = cov(\rho_t, -s_t/(1+s_t)) - E[\rho_t] E[s_t/(1+s_t)] \le 0$ then $x^*=1$ is stochastically locally stable.
\item It is not possible that $E[log(1+z_t)]$ and $E[log(1-z_t/(1+s_t)]$ are both negative, as their sum is positive
\begin{multline}
E[log(1+z_t)] + E[log(1-z_t/(1+s_t)] = \\
E[log(1+z_t) + log(1-z_t/(1+s_t)]= \\
E[log\big((1+z_t)(1-z_t/(1+s_t)\big)]= \\
E[log\big( 1+\rho_t(1-\rho_t)s_t^2/(1+s_t) \big)] > 0,
\end{multline}
and therefore it is not possible that both fixations are stochastically locally stable.
\end{itemize}

\paragraph{Examples.}
First, if $s_t$ and $\rho_t$ are independent ($cov(s_t, \rho_t)=0$) and $s_t$ is symmetric around zero, then $E[-z_t/(1+s_t)] > 0$ (because $E[-z_t/(1+s_t)] = - E[\rho_t] \cdot E[s_t/(1+s_t)]$ and $E[s_t/(1+s_t)] < E[s_t] = 0$).
Therefore, fixation of $B$ is stochastically locally stable and fixation of $A$ almost surely does not occur.
For example, let $s_t \sim U(-1, 1)$ and $\rho_t \sim U(0,1)$ independently (in particular, $cov(\rho_t, s_t)=0$), then $E[log(1+z_t)]\approx -0.07315$ and $E[log(1-z_t/(1+s_t))]\approx 0.2337$.

@@

@@

We obtain a result similar to result 11 from~\citet{Ram2018}:
The mean $\mu(x)$ and variance $\sigma^2(x)$ of the change in the frequency $x$ of phenotype $A$ in the case of a cycling environment $AkBl$, where $k+l=n$, are
\begin{equation} \label{eq:drift_diffusion_terms}
\mu(x) = S_n x(1-x)
\quad \text{and} \quad
\sigma^2(x) = n x (1-x),
\end{equation}
where $S_n = \sum_{t=1}^{n}{z_t}$ and $z_t = \rho_t s_t$.
Furthermore, combining \eqref{eq:drift_diffusion_terms} with eqs.~46-47 from \citet{Ram2018}, we find that the fixation probability of phenotype $A$, starting from a frequency $x$ of phenotype $A$, is
\begin{equation}
u(x) = \frac{1 - e^{-2 \frac{S_n}{n} x}}{1 - e^{-2 \frac{S_n}{n}}}.
\end{equation}
From result 10 in \citep{Ram2018}, $u(x)$ is monotone increasing in $S_n/n$ which is the average selection coefficient of $A$ weighted by the vertical transmission rates $\rho_t$.
Therefore, if $s_t$ and $\rho_t$ are positively (negatively) correlated, $S_n/n$ increases (decreases), and the fixation probability $u(x)$ increases (decreases). 
This occurs because selection only affects those individuals that transmit their phenotype to their own offspring (i.e. vertically), and a fraction $1-\rho_t$ of the population individuals is effectively masked (for better or worse) from selection at each generation.

@@

Figures \ref{fig:migration_rho} and \ref{fig:migration_m} show the stable frequencies of phenotype $A$~ (eq.~\ref{eq:migration_model_unconditional_symmetric_polymorphism}) and the stable population mean fitnesses in the two demes with symmetric selection where $A$ is favored in deme 1 and $B$ is favored in deme 2 with similar selection intensities.
Notably, in the absence of oblique transmission (\autoref{fig:migration_m}, left column), migration decreases the differences between the demes and reduces the population mean fitnesses.
With some oblique transmission, but equal in both demes, results are similar (not shown).
However, if oblique transmission is stronger in deme 2 than in deme 1 (\autoref{fig:migration_m}, middle and right columns), the stable frequency of $A$ increases in both demes.
Therefore, the mean fitness in deme 1 decreases to a lesser extent than in deme 2, and even increases when the migration rate is high enough.

@@
